HEP Software Foundation Community White Paper -- Data Processing Framework WG Report

### 2018-04-23 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Update df_fwk_summary.tex
Sync the google doc to this input source for the final paper.  Everything is one except for the final 3-5 year goals.

The time periods of interest for this document are DUNE and HL-LHC,
which will deliver on the order of 50 PB data events per year per
experiment . The results of the proposed R\&D ought to be used for
building the final software systems that will be utilized in
commissioning and operations of these experiments and the processing
of the data.

@@

\begin{enumerate}
\item
    Changes needed in the programming model that are necessary to
    handle the massive parallelism that will be present throughout all
    layers in the available compute facilities. This is necessary
    because of the ever-increasing availability of specialized compute
    resources, and includes GPGPUs, Tensor Processing Units (TPUs),
    and tiered memory systems integrated with storage, ultra
    high-speed network interconnects.
\item
    Challenges related to advanced detector technology, like finer
    granularity high bandwidth continuous readout DAQ, and the need
    for a very tight feedback loop for conditions which in turn
    requires a focus on low-latency solutions.
\item
    Tighter integration with the computing model: informing
    higher-level systems (workflow, data management, workload
    management, job management) based on fundamental changes in
    compute facilities.
\item
    Support structures that permit frameworks to be collaboratively
    developed and maintained across a large number of
    experiments. This includes excellent integration with the wider
    ecosystem encompasses development, deployment, and runtime
    components.
\item
    Provide flexibility and the interfaces required to efficiently
    integrate the various parallelization efforts on simulation,
    reconstruction and other fields.
\end{enumerate}

@@

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. A complete description
of framework use cases written for the Gaudi collaboration is
described in {[}2.fixme{]} and these are sufficiently general to apply
to all HEP experiments. The Gaudi framework {[}70{]} was originally
developed by LHCb, but is also used by ATLAS and various non-LHC
experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
which was forked to provide the art framework for the Fermilab
Intensity Frontier experiments~\cite{Green:2012gv}.  Belle II uses
basf2~\cite{1742-6596-331-3-032024}. The linear collider community
developed and uses Marlin~\cite{Gaede:2006pj}. The FAIR experiments
use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
and ALICE are now developing a new framework, which is called
O2~\cite{O2}. At the time of writing, most major frameworks support
basic parallelisation, both within and across events, based on a
task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}.

@@

Current practice for throughput-maximizing system architectures have
constrained the scope of framework designs. Framework applications
have largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this
tradition starting in the beginning of Run 2, by utilizing all cores
on one virtual node in one process space using threading. ATLAS is
currently using a multi-process fork-and-copy-on-write solution to
remove the constraint of one core/process. Both experiments were
driven to solve this problem, by the ever growing needs for more
memory per process brought on by the increasing complexity of LHC
events. Current practice manages system-wide (or facility-wide)
scaling by dividing up datasets, generating a framework application
configuration, and scheduling jobs on nodes/cores to consume all
available resources. Given anticipated changes in hardware
(heterogeneity, connectivity, memory, storage) available at large
computing facilities, the interplay between workflow/workload
management systems and framework applications need to be carefully
examined. It may be advantageous to permit framework applications (or
systems) to span resources, permitting them to be first-class
participating in the business of scaling within a facility.

\section{Roadmap}
\label{sec:roadmap}

@@

\paragraph{Concept refinement} Jointly identify key abstractions that
made frameworks good for HEP in detail beyond what can be described in
this paper. Identify and describe where individual frameworks have
similarly or uniquely implemented these concepts. A publishable paper
should come of this that will serve as an agreed-upon guide for where
we can hope to go.

\paragraph{Technology investigations} There are three key areas that
ought to be explored to help determine future direction with regards
to software technology. The areas are: (1) task-based programming
tools, (2) parallel number crunching libraries, and (3) framework
workflow management.

@@

\paragraph{Domain-specific language} We believe it is useful to
describe the language that HEP uses to define and describe how to
simulate and reconstruct physics events. By language, we mean the
terms used to communicate and express how tasks are described and
carried out within a framework. This includes not only expressing data
dependencies, but also resources preferences and constraints, such a
GPU. The goal here is to provide enough information for a group to
take on development of domain-specific libraries components and tools
that will increase the efficiency of carrying out physics.
\ifixme{the wording here needs much improvement. Maybe a good example
  is how ML toolkits have evolved over the past few years - the
  abstractions that have been developed have greatly increased
  productivity and growth in the ML space}

\paragraph{Coupling with ongoing R\&D activities} Forward-looking work
is underway as part of projects funded through government agencies,
laboratories, and collaborations. We want to be sure that relevant
ideas and accomplishment are known, and that the groups doing this
work have a place to report to and receive feedback for everyone's
benefit.

>>>>

The time periods of interest for this document are DUNE and HL-LHC,
which will deliver on the order of 50 PB data events per year per
experiment. The results of the proposed R\&D ought to be used for
building the final software systems that will be utilized in
commissioning and operations of these experiments and the processing
of the data.

@@

\begin{enumerate}
\item
    Changes needed in the programming model that are necessary to
    handle the massive parallelism that will be present throughout all
    layers in the available compute facilities. This is necessary
    because of the ever-increasing availability of specialized compute
    resources, and includes GPGPUs, Tensor Processing Units (TPUs),
    and tiered memory systems integrated with storage, and ultra
    high-speed network interconnects.
\item
    Challenges related to advanced detector technology, like finer
    granularity, high bandwidth continuous readout DAQ, and the need
    for a very tight feedback loop for conditions, which in turn
    requires a focus on low-latency solutions.
\item
    Tighter integration with the computing model: informing
    higher-level systems (workflow, data management, workload
    management, job management) based on fundamental changes in
    compute facilities.
\item
    Support structures that permit frameworks to be collaboratively
    developed and maintained across a large number of
    experiments. This includes excellent integration with the wider
    ecosystem encompasses development, deployment, and runtime
    components.
\item
    Provide flexibility and the interfaces required to efficiently
    integrate the various parallelization efforts on simulation,
    reconstruction and other fields.
\end{enumerate}

@@

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. A complete description
of framework use cases written for the Gaudi collaboration is
described in {[}2.fixme{]} and these are sufficiently general to apply
to all HEP experiments. The Gaudi framework {[}70{]} was originally
developed by LHCb, but is also used by ATLAS and various non-LHC
experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
{[}71{]} which was forked to provide the art framework for the Fermilab
Intensity Frontier experiments~\cite{Green:2012gv},{[}72{]}.  Belle II uses
basf2~\cite{1742-6596-331-3-032024},{[}73{]}. The linear collider community
developed and uses Marlin~\cite{Gaede:2006pj},{[}74{]}. The FAIR experiments
use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
and ALICE are now developing a new framework, which is called
O2~\cite{O2},{[}75{]}. At the time of writing, most major frameworks support
basic parallelisation, both within and across events, based on a
task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}. O2 already 
includes additional multi-node setups and communication.

@@

Current practice for throughput-maximizing system architectures have
constrained the scope of framework designs. Framework applications
have largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this
tradition starting in the beginning of Run 2, by utilizing all cores
on one virtual node in one process space using threading. ATLAS is
currently using a multi-process fork-and-copy-on-write solution to
remove the constraint of one core/process, and is now moving to the 
multithreading approach too. Both experiments were
driven to solve this problem, by the ever growing needs for more
memory per process brought on by the increasing complexity of LHC
events. Current practice manages system-wide (or facility-wide)
scaling by dividing up datasets, generating a framework application
configuration, and scheduling jobs on nodes/cores to consume all
available resources. Given anticipated changes in hardware
(heterogeneity, connectivity, memory, storage) available at large
computing facilities, the interplay between workflow/workload
management systems and framework applications need to be carefully
examined. It may be advantageous to permit framework applications (or
systems) to span resources, permitting them to be first-class
participatents in the business of scaling within a facility. O2 provides
a successful proof-of-principle of this approach.

\section{Roadmap}
\label{sec:roadmap}
Forward-looking work is underway as part of projects funded through government agencies, laboratories, and collaborations.  We want to be sure that relevant ideas and accomplishment are known, and that the groups doing this work have a place to report to and receive feedback for everyone’s benefit. 
To organize the community, one needs to establish regular working group meetings, on a bi-monthly basis as we did with the concurrency forum. Face to face workshops after at least the 1st and the 3rd year can be co-hosted with events like CHEP and/or the WLCG workshops. A future planning workshop for transforming the results of the R&D activities into a full development and deployment project plan should happen at the 5-year timescale.

@@

\paragraph{Concept refinement} Jointly identify key abstractions that
made frameworks good for HEP in detail beyond what can be described in
this paper. Identify and describe where individual frameworks have
similarly or uniquely implemented these concepts. It is important to 
describe how these choices are connected to the concrete use-cases. A 
publishable paper should come of this that will serve as an agreed-upon 
guide for where we can hope to go.

\paragraph{Technology investigations} There are four key areas that
ought to be explored to help determine future direction with regards
to software technology. The areas are: (1) task-based programming tools,
(2) inter-process and inter-node communication tools, (3) parallel number 
crunching libraries, and (4) framework workflow management.

@@

\paragraph{Domain-specific language} We believe it is useful to
describe the language that HEP uses to define and describe how to
simulate and reconstruct physics events. By language, we mean the
terms used to communicate and express how tasks are described and
carried out within a framework. This includes not only expressing data
dependencies, but also resource preferences and constraints, such as
GPU. The goal here is to provide enough information for a group to
take on development of domain-specific libraries components and tools
that will increase the efficiency of carrying out physics. A good example
is how ML toolkits have evolved over the past few years. The 
abstractions that have been developed have greatly increased productivity 
and growth in the ML space such as the abstractions in Tensor Flow that 
allow a coding of the matrix algebra that then gets remapped internally to 
match the shape of the data being operated on.  The user only has take care
of getting the domain science functions correct.
